Extreme Digitisation For Ground-Based Cosmic Microwave Background Experiments

### 2019-05-17 response letter/response_letter.tex

%\title{MNRAS: MN-18-3671-MJ\\Response to the Referee}

%\maketitle

\begin{center}
\Large MNRAS: MN-18-3671-MJ\\Response to the Referee
\end{center}

Thank you for taking the time to reassess our manuscript and review our response. We appreciate your comments with regards to the compression power of few-bit digitisation in conjunction with existing data reduction methods. Please find a detailed response to the points you raised below.

@@

\hrule
\vspace{0.1cm}
\hrule

\vspace{0.5cm}

Reviewer's comment:

\indent \question{I am a bit shocked that the performance in conjunction with LPC-/PCM-based compression algorithms is as good as claimed here [...] My suspicion here is that the synthetic data under test have much higher entropy [...]}

@@

\vspace{0.25cm}

Authors' reply:

\indent We appreciate the concern raised. To address this issue appropriately we followed the referee's suggestion and repeated the analysis using genuine experimental data.
We obtain SPT-3G data and use the FLAC compression algorithm used by the SPT. We apply few-bit digitisation to the TOD and apply FLAC. This allows us to compare the resulting file sizes between the few-bit data and the typical compression rates achieved for SPT-3G data.
Applying FLAC to few-bit timestreams does not lead to additional data volume reduction. We altered subsection 3.3 to reflect this insight.

@@

\vspace{0.25cm}

Changes to the manuscript:

\indent \textbf{\changed{3.3 Existing Compression Techniques}}

\changed{Current-generation ground-based CMB experiments employ a combination of lossly, e.g. downsampling, and lossless compression, e.g. FLAC, bzip, LZMA, to manage their transmission bottlenecks. We test whether the data volume of timestreams that have undergone few-bit digitisation can be reduced further by applying the most popular lossless compression algorithm: FLAC. }

\changed{We obtain 2 minutes of SPT-3G TOD and apply the digitisation schemes in equations \ref{eq:1bit}-\ref{eq:3bit} before compressing the data further using FLAC. We compare the file size of FLAC compressed few-bit timestreams to the typical data reduction rate achieved for 24-bit SPT-3G data.}

\changed{We find that using FLAC in conjunction with extreme digitisation does not yield additional data volume reduction. For 3-bit digitisation, both the FLAC and raw timestream have $\sim$3 bits per data point. In fact, due to the overhead required by FLAC compressed files perform marginally worse than uncompressed few-bit streams. This is unsurprising, given that we have already maximally compressed the data using extreme digitisation. For comparison, FLAC reduces the full 24-bit  SPT-3G data to $6-8$ bits of entropy per number. For this real-world case, three-bit digitisation reduces the data volume by a factor of $2-3$.}

@@

\hrule

\vspace{0.5cm}

Reviewer's comment:

\indent \question{[...] what is the impact on non-linearity in power spectrum estimation? [...]}

@@

\vspace{0.25cm}

Authors' reply:

\indent Thank you for raising this point. Indeed the non-linearity introduced by few-bit digitisation disappears in the power spectrum estimation. This is because many TOD points contribute to each map pixel and hence the central limit theorem ensures that the distribution within each pixel approaches a Gaussian. We appended a sentence stating this to subsection 2.4 (Power Spectrum Estimation).

@@

\vspace{0.25cm}

Changes to the manuscript:

\indent \changed{Non-linearity introduced by the digitisation process disappears in the final map due to the large number of TOD points that contribute to each map pixel. Thus the digitisation effects are not an issue for power spectrum estimation.}

### 2019-05-17 digitisation.tex
Add files via upload
Updated for second response.

The large size of the time ordered data of cosmic microwave background experiments presents challenges for mission planning and data analysis. These issues are particularly significant for Antarctica- and space-based experiments, which depend on satellite links to transmit data. We explore the viability of reducing the time ordered data to few bit numbers to address these challenges. Unlike lossless compression, few bit digitisation introduces additional noise into the data. We present a set of one, two, and three bit digitisation schemes and measure \changed{the} increase in noise in the cosmic microwave background temperature and polarisation power spectra. The digitisation noise is independent of angular scale and is well-described as a constant percentage of the original detector noise. Three bit digitisation increases the map noise level by $< 2\%$, while reducing the data volume by a factor of \changed{twenty relative to 64}-bit floats. \changed{Comparable gains are seen when extreme digitisation is combined with  
established lossless compression algorithms. Extreme} digitisation is a promising strategy for upcoming experiments.

@@

The Planck mission has already demonstrated the success of extreme digitisation for space-based CMB \changed{experiments} \citep{maris2003}. \cite{jenet1998} explored the application of few bit digitisation to radio pulsar timing measurements. Recently Clearwater et al. (private communication) have demonstrated the advantages of using one and two bit data when searching for continuous gravitational waves using the Laser Interferometer Gravitational-Wave Observatory (LIGO).

@@

Digitisation discretises an input signal by sorting it into $N$ ranges, such that an input between $x_i$ and $x_{i+1}$ produces an output value $y_i$. The set of parameters $N, x_i, y_i$ fully specify a digitisation scheme. Conventionally one chooses $x_{1} = -\infty$ and $x_{N+1} = \infty$, i.e. values beyond some threshold saturate and yield identical output. In order to quantify the performance of a given digitisation scheme we define the distortion as
\begin{equation}\label{eq:distdef}
D = \left\langle  \left( s - \hat{s} \right)^2 \right\rangle,
\end{equation}
where $s$ is the input and $\hat{s}$ the output signal. For a stochastic input signal we can calculate an amplitude probability distribution $p(x)$. \changed{This allows us to re-express the distortion as a sum over the digitisation levels:}

@@

Since we wish to minimise the distortion, we set \changed{its derivatives} with respect to $x_i$ and $y_i$ to zero. Starting with $x_i$ we obtain

@@

\begin{equation} \label{eq:digitareacondition}
\int_{x_i}^{x_{i+1}} \left( x-y_i \right) p(x) dx = 0.
\end{equation}
This \changed{condition} implies that we should choose $y_i$, such that it halves the area underneath $p(x)$ in the interval from $x_i$ to $x_{i+1}$.

To progress further, we must consider the probability distribution $p(x)$ of the input signal. 
\changed{In this work, we are interested in the compression of the TOD from each detector of a ground-based CMB experiment. 
Thus we can safely assume that the input signal is in the low signal-to-noise regime. 
Furthermore, we take the simplest noise profile: Gaussian, white detector noise\footnote{\changed{As discussed in the appendix, this treatment also applies to more realistic noise profiles, because the large number of points lying in each map pixel allows the application of the central limit theorem.}}. 
Together these assumptions imply that the input, $x$, is normally distributed, with} $p(x) = (1/\sqrt{2\pi\sigma^2}) e^{-x^2/2\sigma^2}$.

\changed{\citet{max1960} solved Equations \ref{eq:digitequalspacecondition} and \ref{eq:digitareacondition} for 1-, 2- and 3-bit digitisation for this choice of probability distribution, $p(x)$. 
We quote their results below and refer the reader to the aforementioned work for the derivation.  
As one would expect given the symmetry in $p(x)$, the 1-bit (N=2 output levels) case is simply:
\begin{equation} \label{eq:1bit}
\hat{s}_1(t) = \left\{ \begin{array}{lr}
1, & s(t) > 0\\
-1, & s(t) \leq 0
\end{array} \right. \hspace{0.5cm}.  \end{equation}
Essentially, the optimal 1-bit digitisation scheme applies the sign function to the TOD. 
}

\changed{
For two bit digitisation, we can encode $N=4$ output levels. The optimal set of thresholds and digitisation levels is given by the four-level function:}
\begin{equation}  \label{eq:2bit}
\hat{s}_2(t) = \left\{ \begin{array}{rl}
1.51 \sigma, & s(t) \geq 0.98 \sigma\\
0.45 \sigma, & 0 \leq s(t) < 0.98 \sigma\\
-0.45 \sigma, & -0.98 \sigma \leq s(t) < 0\\
-1.51 \sigma, & -0.98 \sigma < s(t)\\
\end{array} \right. \hspace{0.5cm}.  \end{equation}
%C: should i round to equal significant figures here? Unfortunately these are the exact values that i used...

\changed{Finally, with 3 bits we may specify $N=8$ output levels. The optimal three bit digitisation is described by the eight-level function:}
\begin{equation}  \label{eq:3bit}
\hat{s}_3(t) = \left\{ \begin{array}{rl}
2.15 \sigma, & s(t) \geq 1.75 \sigma\\
1.34 \sigma, & 1.05 \sigma \leq s(t) < 1.75 \sigma\\
0.76 \sigma, & 0.50 \sigma \leq s(t) < 1.05 \sigma\\
0.25 \sigma, & 0 \leq s(t) < 0.50 \sigma\\
-0.25 \sigma, & -0.50 \sigma \leq s(t) < 0\\
-0.76 \sigma, & -1.05 \sigma \leq s(t) < -0.50 \sigma\\
-1.34 \sigma, & -1.75 \sigma \leq s(t) < -1.05 \sigma\\
-2.15 \sigma, & -1.75 \sigma < s(t)\\
\end{array} \right. \hspace{0.5cm}.  \end{equation}

\changed{The digitisation schemes above are designed to minimise the distortion of the TOD as defined in Equation \ref{eq:distdef}. This does not ensure that the digitised output has the same signal power as the input. However, the output of the schemes given above can be rescaled by a constant to achieve power conservation. This is demonstrated in the appendix, where we also derive an analytical expression for the normalisation constant. We use this analytical rescaling for all results in this work.
}

@@

We rebin the calculated power spectra as mentioned \changed{when applicable} and assign Fisher matrix error bars \citep{tegmarkdesignersguide1997}.

As seen in Figure \ref{fig:psrecover} (appendix) all spectra of extremely digitised data recover the input to a satisfying degree. Differences between few bit and the input power spectra are due to observing a single CMB realisation and residual boundary effects.

@@

The deduced additional noise for one, two, and three bit digitisation is shown in Table \ref{tab:extranoisewhite} in the appendix. We see that 3 bit digitisation performs best, followed by two bit and one bit. This is expected, since more bits allow a more faithful representation of the input signal. We observe that the fractional increase, $\Delta \sigma / \sigma$, is independent of the hits per pixel in the maps and that the compression performs as well for polarisation as is does for temperature data. \changed{Finally, it is striking how little noise is added. On average one-bit digitisation leads to a $(25.13\pm 0.94)\%$ increase in the map noise level, two-bit yields a $(6.40\pm0.41)\%$ increase and three-bit adds only $(1.76\pm0.22)\%$ ($1\sigma$ confidence intervals given). This is impressive, given that all schemes considered reduce the TOD volume by at least an order of magnitude.}

@@

\begin{figure}[htb]\centering
\includegraphics[width=0.5\textwidth,clip]{diffwn.pdf}
  \caption[Current ]{
  \changed{(FIGURE HAS BEEN CHANGED) TT, EE, and BB power spectra (upper, middle and lowest panel respectively) of difference maps for one-bit (green marker, high), two-bit (orange marker, medium), and three-bit digitisation (violet marker, low). The control case using 64-bit TOD is also shown (black marker). All power spectra are rebined to $\Delta \ell = 200$. The vertical axes are logarithmic. The noise increase due to digitisation appears to be independent of angular scale. This plot shows results for a simulation with $1,024,000$ hits per map pixel using Gaussian white detector noise.}
\label{fig:diffpswn}
}
\end{figure}

@@

While digitisation clearly works extremely well for white noise, real ground-based CMB experiments face several sources of low-frequency noise. \changed{High on the list is atmospheric noise. Modelling the atmosphere as a thin sheet dominated by Kolmogorov turbulence, we can describe fluctuations with the noise profile \citep{lay2000}:}

\changed{
\begin{equation} \label{eq:1/fnoiseform}
\left| N_{1/f}(\ell_{\mathrm{scan}}) \right|^2 = \left| N_{\mathrm{white}} \right|^2 \left[ 1 + \left( \frac{\ell_{\mathrm{scan}}}{\ell_0} \right)^{-3/2} \right]^2,
\end{equation}
}

@@

where $N_{\mathrm{white}}$ denotes white detector noise. In this expression \changed{$\ell_{\mathrm{scan}}$} describes the multipole moment of angular features of corresponding size along the scan direction. The scale at which such modes are amplified appreciably by the second term in equation \ref{eq:1/fnoiseform} is set by \changed{$\ell_0$}. \changed{The exponent of $\ell_{\mathrm{scan}}/\ell_0$ is chosen to model observation at a constant, moderate elevation \citep{lay1997}. In line with the observed noise profile of the SPT we choose $\ell_0 = 1000$ for temperature and $\ell_0 = 0$ for polarisation; this means that power is doubled at $\lknee \approx 1800$ and $\lknee \approx 90$ for temperature and polarisation, respectively \citep{henning2018}.} In practice CMB experiments suppress power in the lowest modes through a range of techniques, including fitting polynomials, sines, and cosines to the TOD. We simulate successful removal of these modes by applying a high pass filter to the noise, such that

\changed{
\begin{equation} \label{eq:1/fnoiseformfinal}
\left| N (\ell_{\mathrm{scan}}) \right|^2 = \left| N_{1/f}(\ell_{\mathrm{scan}}) \right|^2 \exp\left[ -2\left( \frac{\lcut}{\ell_{\mathrm{scan}}} \right)^{6} \right],
\end{equation}
}

@@

We use the same digitisation schemes as for white noise, but \changed{replace} the parameter $\sigma$ in equations \ref{eq:1bit}, \ref{eq:2bit} and \ref{eq:3bit} by considering Parseval's theorem on the above noise profile. The appropriate value is obtained by \changed{numerically} evaluating the integral

@@

We recover similar results to the white noise case, with three bit digitisation performing the best, followed by two bit, and one bit. Table \ref{tab:extranoiseoof} in the appendix displays the fractional increase to the detector noise level due to digitisation, $\Delta \sigma / \sigma$. While polarisation observations seem to incur a slightly higher noise penalty compared to temperature, deviations do not exceed $3\sigma$. \changed{On average one-bit digitisation leads to a $(24.4\pm 1.4)\%$ increase in the map noise level, two-bit adds $(6.21\pm0.50)\%$, and three-bit yields an increase of only $(1.71\pm0.21)\%$ ($1\sigma$ confidence intervals given). Thus the effectiveness of few bit digitisation is not limited to the white noise case, but \changed{extends} to more realistic noise profiles.}

@@

\subsection{\changed{Existing Compression Techniques}}

\changed{Current-generation ground-based CMB experiments employ lossless and lossy compression techniques to manage their transmission bottlenecks. To test whether few-bit digitisation is competitive to existing lossy compression techniques, it is necessary to see if it still provides a substantial reduction in data volume if used in conjunction with lossless compression algorithms.}

\changed{We focus on lossless compression via the Lempel-Ziv-Markoc chain algorithm (LZMA), which is used by the Background Imaging of Cosmic Extragalactic Polarization (BICEP) experiment, and the free lossless audio codec (FLAC), which the SPT and POLARBEAR experiments employ (private communication). We examine the compression of a single CES of our simulation, consisting of ca. $10^7$ data points, by applying LZMA and FLAC separately to four timestreams: the original 64-bit TOD, and copies that have undergone 1-, 2-, and 3-bit digitisation.}

\changed{We find that even though the data reduction power of few-bit digitisation is decreased when used in conjunction with lossless compression algorithms, substantial savings are still achieved. Applying few-bit digitisation to the TOD prior to compression with FLAC reduces the final file size by a factor $5.89$ compared to the FLAC compressed 64-bit data. This is the same for all digitisation schemes considered. Working with LZMA, the data volume of the TOD can be reduced with respect to compressed 64-bit data by factors of $17.34$, $26.23$, and $44.48$ for 3-, 2-, and 1-bit digitisation, respectively. Few-bit digitisation still achieves considerable data reduction if used together with established lossless compression algorithms.}

@@

We present a set of one, two, and three bit digitisation schemes. For white and $1/f$ detector noise alike, we find that optimal three bit digitisation adds as little as $<2\%$ to the map noise level. This is true for temperature and polarisation observations. No change in the results is observed for maps \changed{with different numbers of} hits per pixel and no angular scale dependence was observed in the added noise. \changed{Combining few-bit digitisation with currently used lossless compression codecs reduces its compression power only slightly. Applying three-bit digitisation to TOD prior to lossless compression via FLAC or LZMA reduces the data volume by a further factor of $5.89$ or $17.34$, respectively.}

@@

We are grateful for insightful discussions about the prospects of extreme digitisation for CMB experiments with Andrew Melatos and Patrick Clearwater.
We thank the \changed{referee as well as Nikhel Gupta, Srinivasan Raghunathan, Federico Bianchini, Andrew Melatos, and Patrick Clearwater for valuable feedback on the manuscript.}
We acknowledge support from an Australian Research Council Future Fellowship (FT150100074), and also from the University of Melbourne. 
This research used resources of the National Energy Research Scientific Computing Center, which is supported by the Office of Science of the U.S. Department of Energy under Contract No. DE-AC02-05CH11231. 
We acknowledge the use of the Legacy Archive for Microwave Background Data Analysis (LAMBDA). Support for LAMBDA is provided by the NASA Office of Space Science.

@@

\begin{figure*}[h!t]\centering
\vspace{-1.5cm}
\includegraphics[width=0.8\textwidth,clip]{psrecovery.pdf}
  \caption[Current ]{
  \changed{(FIGURE HAS BEEN CHANGED) Input power spectra used to generate the CMB template map plus detector noise level (black line) and digitisation noise (grey line); power spectra using three- (violet marker, right), two- (orange marker, middle), and one-bit (green marker, left) TOD. Power spectra are rescaled to $D_\ell = \ell (\ell+1) C_\ell /2\pi$ and bined to $\Delta \ell = 100$. Main panels show the TT (top), EE (middle), and BB (bottom) power spectra; accompanying panels show the difference between the input plus noise curves and the power spectra recovered from few-bit TOD. Spectra constructed from few-bit TOD recover the input well. This plot shows results for a simulation with $1,024,000$ hits per map pixel using Gaussian white detector noise.}
\label{fig:psrecover}
}
\end{figure*}

@@

As mentioned in subsection \S\ref{subsec:extremedigitisation}, equation \ref{eq:distdef} does not demand \changed{that a digitisation scheme conserves power. However, it} is relatively simple to impose power conservation, since it boils down to rescaling the digitised output by a constant, $\gamma$. We derive an expression for $\gamma$, such that a cross-spectrum of two digitised maps yields an unbiased estimate of the input power spectrum. Mathematically, this is equivalent to:

@@

\changed{ where we sum over all output levels of both timestreams $(1, 2)$ and $P$ denotes the probability of an output level occurring. For current-generation CMB experiments it is common to draw from $>10^5$ samples to obtain an estimate of a map pixel. This corresponds to adding many noise realisation on top of the same signal, $\mu$. 
The central limit theorem ensures that the amplitude probability distribution function of the noise in each pixel is Gaussian. 
The probabilities $P$ in the above expression can therefore be expressed as integrals over normal distributions, $p(\xi)$, with mean zero and standard deviation $\sigma$.}

@@

\begin{equation}
\begin{aligned}
&= \left\{ \sum_{i=0}^N  \frac{y_i}{\sqrt{pi}} \left[ \sum_{n = 0}^\infty \frac{(-1)^n}{n! (2n+1)} \frac{1}{(\sqrt{2}\sigma)^{2n+1}} \left( \sum_{k=0}^{2n+1} {2n+1 \choose k} \mu^k ( x_{i+1}^{2n+1-k} - x_{i}^{2n+1-k} ) \right) \right] \right\} \\
&\times \left\{ \sum_{j=0}^N  \frac{y_j}{\sqrt{2}} \left[ \sum_{m = 0}^\infty \frac{(-1)^m}{m! (2m+1)} \frac{1}{(\sqrt{2}\sigma)^{2m+1}} \left( \sum_{l=0}^{2m+1} {2m+1 \choose l} \mu^l ( x_{j+1}^{2m+1-l} - x_{j}^{2m+1-l} ) \right) \right] \right\} \changed{,}
\end{aligned}
\end{equation}

\changed{where} we have used the Maclaurin series expansion of the error function and the binomial expansion. Moving the $i, j$ summation forwards yields

@@

where $c$ is some constants that \changed{depends} on the specific digitisation scheme chosen. Returning to equation \ref{eq:normcrosspower} we see

>>>>

The large size of the time ordered data of cosmic microwave background experiments presents challenges for mission planning and data analysis. These issues are particularly significant for Antarctica- and space-based experiments, which depend on satellite links to transmit data. We explore the viability of reducing the time ordered data to few bit numbers to address these challenges. Unlike lossless compression, few bit digitisation introduces additional noise into the data. We present a set of one, two, and three bit digitisation schemes and measure the increase in noise in the cosmic microwave background temperature and polarisation power spectra. The digitisation noise is independent of angular scale and is well-described as a constant percentage of the original detector noise. Three bit digitisation increases the map noise level by $< 2\%$, while reducing the data volume by a factor of \changed{ten relative to 32-bit floats. Lossless compression of the digitised data does not lead to additional data volume reduction.} Extreme digitisation is a promising strategy for upcoming experiments.

@@

The Planck mission has already demonstrated the success of extreme digitisation for space-based CMB experiments \citep{maris2003}. \cite{jenet1998} explored the application of few bit digitisation to radio pulsar timing measurements. Recently Clearwater et al. (private communication) have demonstrated the advantages of using one and two bit data when searching for continuous gravitational waves using the Laser Interferometer Gravitational-Wave Observatory (LIGO).

@@

Digitisation discretises an input signal by sorting it into $N$ ranges, such that an input between $x_i$ and $x_{i+1}$ produces an output value $y_i$. The set of parameters $N, x_i, y_i$ fully specify a digitisation scheme. Conventionally one chooses $x_{1} = -\infty$ and $x_{N+1} = \infty$, i.e. values beyond some threshold saturate and yield identical output. In order to quantify the performance of a given digitisation scheme we define the distortion as
\begin{equation}\label{eq:distdef}
D = \left\langle  \left( s - \hat{s} \right)^2 \right\rangle,
\end{equation}
where $s$ is the input and $\hat{s}$ the output signal. For a stochastic input signal we can calculate an amplitude probability distribution $p(x)$. This allows us to re-express the distortion as a sum over the digitisation levels:

@@

Since we wish to minimise the distortion, we set its derivatives with respect to $x_i$ and $y_i$ to zero. Starting with $x_i$ we obtain

@@

\begin{equation} \label{eq:digitareacondition}
\int_{x_i}^{x_{i+1}} \left( x-y_i \right) p(x) dx = 0.
\end{equation}
This condition implies that we should choose $y_i$, such that it halves the area underneath $p(x)$ in the interval from $x_i$ to $x_{i+1}$.

To progress further, we must consider the probability distribution $p(x)$ of the input signal. 
In this work, we are interested in the compression of the TOD from each detector of a ground-based CMB experiment. 
Thus we can safely assume that the input signal is in the low signal-to-noise regime. 
Furthermore, we take the simplest noise profile: Gaussian, white detector noise\footnote{As discussed in the appendix, this treatment also applies to more realistic noise profiles, because the large number of points lying in each map pixel allows the application of the central limit theorem.}. 
Together these assumptions imply that the input, $x$, is normally distributed, with $p(x) = (1/\sqrt{2\pi\sigma^2}) e^{-x^2/2\sigma^2}$.

\citet{max1960} solved Equations \ref{eq:digitequalspacecondition} and \ref{eq:digitareacondition} for 1-, 2- and 3-bit digitisation for this choice of probability distribution, $p(x)$. 
We quote their results below and refer the reader to the aforementioned work for the derivation.  
As one would expect given the symmetry in $p(x)$, the 1-bit (N=2 output levels) case is simply:
\begin{equation} \label{eq:1bit}
\hat{s}_1(t) = \left\{ \begin{array}{lr}
1, & s(t) > 0\\
-1, & s(t) \leq 0
\end{array} \right. \hspace{0.5cm}.  \end{equation}
Essentially, the optimal 1-bit digitisation scheme applies the sign function to the TOD.

For two bit digitisation, we can encode $N=4$ output levels. The optimal set of thresholds and digitisation levels is given by the four-level function:
\begin{equation}  \label{eq:2bit}
\hat{s}_2(t) = \left\{ \begin{array}{rl}
1.51 \sigma, & s(t) \geq 0.98 \sigma\\
0.45 \sigma, & 0 \leq s(t) < 0.98 \sigma\\
-0.45 \sigma, & -0.98 \sigma \leq s(t) < 0\\
-1.51 \sigma, & -0.98 \sigma < s(t)\\
\end{array} \right. \hspace{0.5cm}.  \end{equation}
%C: should i round to equal significant figures here? Unfortunately these are the exact values that i used...

Finally, with 3 bits we may specify $N=8$ output levels. The optimal three bit digitisation is described by the eight-level function:
\begin{equation}  \label{eq:3bit}
\hat{s}_3(t) = \left\{ \begin{array}{rl}
2.15 \sigma, & s(t) \geq 1.75 \sigma\\
1.34 \sigma, & 1.05 \sigma \leq s(t) < 1.75 \sigma\\
0.76 \sigma, & 0.50 \sigma \leq s(t) < 1.05 \sigma\\
0.25 \sigma, & 0 \leq s(t) < 0.50 \sigma\\
-0.25 \sigma, & -0.50 \sigma \leq s(t) < 0\\
-0.76 \sigma, & -1.05 \sigma \leq s(t) < -0.50 \sigma\\
-1.34 \sigma, & -1.75 \sigma \leq s(t) < -1.05 \sigma\\
-2.15 \sigma, & -1.75 \sigma < s(t)\\
\end{array} \right. \hspace{0.5cm}.  \end{equation}

The digitisation schemes above are designed to minimise the distortion of the TOD as defined in Equation \ref{eq:distdef}. This does not ensure that the digitised output has the same signal power as the input. However, the output of the schemes given above can be rescaled by a constant to achieve power conservation. This is demonstrated in the appendix, where we also derive an analytical expression for the normalisation constant. We use this analytical rescaling for all results in this work.

@@

We rebin the calculated power spectra as mentioned when applicable and assign Fisher matrix error bars \citep{tegmarkdesignersguide1997}.

As seen in Figure \ref{fig:psrecover} (appendix) all spectra of extremely digitised data recover the input to a satisfying degree. Differences between few bit and the input power spectra are due to observing a single CMB realisation and residual boundary effects. \changed{Non-linearity introduced by the digitisation process disappears in the final map due to the large number of TOD points that contribute to each map pixel. Thus the digitisation effects are not an issue for power spectrum estimation.}

@@

The deduced additional noise for one, two, and three bit digitisation is shown in Table \ref{tab:extranoisewhite} in the appendix. We see that 3 bit digitisation performs best, followed by two bit and one bit. This is expected, since more bits allow a more faithful representation of the input signal. We observe that the fractional increase, $\Delta \sigma / \sigma$, is independent of the hits per pixel in the maps and that the compression performs as well for polarisation as is does for temperature data. Finally, it is striking how little noise is added. On average one-bit digitisation leads to a $(25.13\pm 0.94)\%$ increase in the map noise level, two-bit yields a $(6.40\pm0.41)\%$ increase and three-bit adds only $(1.76\pm0.22)\%$ ($1\sigma$ confidence intervals given. This is impressive, given that all schemes considered reduce the TOD volume by at least an order of magnitude.

@@

\begin{figure}[htb]\centering
\includegraphics[width=0.5\textwidth,clip]{diffwn.pdf}
  \caption[Current ]{
  TT, EE, and BB power spectra (upper, middle and lowest panel respectively) of difference maps for one-bit (green marker, high), two-bit (orange marker, medium), and three-bit digitisation (violet marker, low). The control case using 64-bit TOD is also shown (black marker). All power spectra are rebined to $\Delta \ell = 200$. The vertical axes are logarithmic. The noise increase due to digitisation appears to be independent of angular scale. This plot shows results for a simulation with $1,024,000$ hits per map pixel using Gaussian white detector noise.
\label{fig:diffpswn}
}
\end{figure}

@@

While digitisation clearly works extremely well for white noise, real ground-based CMB experiments face several sources of low-frequency noise. High on the list is atmospheric noise. Modelling the atmosphere as a thin sheet dominated by Kolmogorov turbulence, we can describe fluctuations with the noise profile \citep{lay2000}:

\begin{equation} \label{eq:1/fnoiseform}
\left| N_{1/f}(\ell_{\mathrm{scan}}) \right|^2 = \left| N_{\mathrm{white}} \right|^2 \left[ 1 + \left( \frac{\ell_{\mathrm{scan}}}{\ell_0} \right)^{-3/2} \right]^2,
\end{equation}

@@

where $N_{\mathrm{white}}$ denotes white detector noise. In this expression $\ell_{\mathrm{scan}}$ describes the multipole moment of angular features of corresponding size along the scan direction. The scale at which such modes are amplified appreciably by the second term in equation \ref{eq:1/fnoiseform} is set by $\ell_0$. The exponent of $\ell_{\mathrm{scan}}/\ell_0$ is chosen to model observation at a constant, moderate elevation \citep{lay1997}. In line with the observed noise profile of the SPT we choose $\ell_0 = 1000$ for temperature and $\ell_0 = 0$ for polarisation; this means that power is doubled at $\lknee \approx 1800$ and $\lknee \approx 90$ for temperature and polarisation, respectively \citep{henning2018}. In practice CMB experiments suppress power in the lowest modes through a range of techniques, including fitting polynomials, sines, and cosines to the TOD. We simulate successful removal of these modes by applying a high pass filter to the noise, such that

\begin{equation} \label{eq:1/fnoiseformfinal}
\left| N (\ell_{\mathrm{scan}}) \right|^2 = \left| N_{1/f}(\ell_{\mathrm{scan}}) \right|^2 \exp\left[ -2\left( \frac{\lcut}{\ell_{\mathrm{scan}}} \right)^{6} \right],
\end{equation}

@@

We use the same digitisation schemes as for white noise, but replace the parameter $\sigma$ in equations \ref{eq:1bit}, \ref{eq:2bit} and \ref{eq:3bit} by considering Parseval's theorem on the above noise profile. The appropriate value is obtained by numerically evaluating the integral

@@

We recover similar results to the white noise case, with three bit digitisation performing the best, followed by two bit, and one bit. Table \ref{tab:extranoiseoof} in the appendix displays the fractional increase to the detector noise level due to digitisation, $\Delta \sigma / \sigma$. While polarisation observations seem to incur a slightly higher noise penalty compared to temperature, deviations do not exceed $3\sigma$. On average one-bit digitisation leads to a $(24.4\pm 1.4)\%$ increase in the map noise level, two-bit adds $(6.21\pm0.50)\%$, and three-bit yields an increase of only $(1.71\pm0.21)\%$ ($1\sigma$ confidence intervals given). Thus the effectiveness of few bit digitisation is not limited to the white noise case, but extends to more realistic noise profiles.

@@

\subsection{Existing Compression Techniques}

\changed{Current-generation ground-based CMB experiments employ a combination of lossly, e.g. downsampling, and lossless compression, e.g. FLAC, bzip, LZMA, to manage their transmission bottlenecks. We test whether the data volume of timestreams that have undergone few-bit digitisation can be reduced further by applying the most popular lossless compression algorithm: FLAC. }

\changed{We obtain 2 minutes of SPT-3G TOD and apply the digitisation schemes in equations \ref{eq:1bit}-\ref{eq:3bit} before compressing the data further using FLAC. We compare the file size of FLAC compressed few-bit timestreams to the typical data reduction rate achieved for 24-bit SPT-3G data.}

\changed{We find that using FLAC in conjunction with extreme digitisation does not yield additional data volume reduction. For 3-bit digitisation, both the FLAC and raw timestream have $\sim$3 bits per data point. In fact, due to the overhead required by FLAC compressed files perform marginally worse than uncompressed few-bit streams. This is unsurprising, given that we have already maximally compressed the data using extreme digitisation. For comparison, FLAC reduces the full 24-bit  SPT-3G data to $6-8$ bits of entropy per number. For this real-world case, three-bit digitisation reduces the data volume by a factor of $2-3$.}

@@

We present a set of one, two, and three bit digitisation schemes. For white and $1/f$ detector noise alike, we find that optimal three bit digitisation adds as little as $<2\%$ to the map noise level. This is true for temperature and polarisation observations. No change in the results is observed for maps with different numbers of hits per pixel and no angular scale dependence was observed in the added noise. \changed{Combining few-bit digitisation with FLAC compression does not yield additional data volume reduction.}

@@

\changed{We are indebted to Nathan Whitehorn for the details about SPT data compression provided by him.}
We are grateful for insightful discussions about the prospects of extreme digitisation for CMB experiments with Andrew Melatos and Patrick Clearwater.
We thank the referee as well as Nikhel Gupta, Srinivasan Raghunathan, Federico Bianchini, Andrew Melatos, and Patrick Clearwater for valuable feedback on the manuscript.
We acknowledge support from an Australian Research Council Future Fellowship (FT150100074), and also from the University of Melbourne. 
This research used resources of the National Energy Research Scientific Computing Center, which is supported by the Office of Science of the U.S. Department of Energy under Contract No. DE-AC02-05CH11231. 
We acknowledge the use of the Legacy Archive for Microwave Background Data Analysis (LAMBDA). Support for LAMBDA is provided by the NASA Office of Space Science.

@@

\begin{figure*}[h!t]\centering
\vspace{-1.5cm}
\includegraphics[width=0.8\textwidth,clip]{psrecovery.pdf}
  \caption[Current ]{
  Input power spectra used to generate the CMB template map plus detector noise level (black line) and digitisation noise (grey line); power spectra using three- (violet marker, right), two- (orange marker, middle), and one-bit (green marker, left) TOD. Power spectra are rescaled to $D_\ell = \ell (\ell+1) C_\ell /2\pi$ and bined to $\Delta \ell = 100$. Main panels show the TT (top), EE (middle), and BB (bottom) power spectra; accompanying panels show the difference between the input plus noise curves and the power spectra recovered from few-bit TOD. Spectra constructed from few-bit TOD recover the input well. This plot shows results for a simulation with $1,024,000$ hits per map pixel using Gaussian white detector noise.}
\label{fig:psrecover}

\end{figure*}

@@

As mentioned in subsection \S\ref{subsec:extremedigitisation}, equation \ref{eq:distdef} does not demand that a digitisation scheme conserves power. However, it is relatively simple to impose power conservation, since it boils down to rescaling the digitised output by a constant, $\gamma$. We derive an expression for $\gamma$, such that a cross-spectrum of two digitised maps yields an unbiased estimate of the input power spectrum. Mathematically, this is equivalent to:

@@

where we sum over all output levels of both timestreams $(1, 2)$ and $P$ denotes the probability of an output level occurring. For current-generation CMB experiments it is common to draw from $>10^5$ samples to obtain an estimate of a map pixel. This corresponds to adding many noise realisation on top of the same signal, $\mu$. 
The central limit theorem ensures that the amplitude probability distribution function of the noise in each pixel is Gaussian. 
The probabilities $P$ in the above expression can therefore be expressed as integrals over normal distributions, $p(\xi)$, with mean zero and standard deviation $\sigma$.

@@

\begin{equation}
\begin{aligned}
&= \left\{ \sum_{i=0}^N  \frac{y_i}{\sqrt{pi}} \left[ \sum_{n = 0}^\infty \frac{(-1)^n}{n! (2n+1)} \frac{1}{(\sqrt{2}\sigma)^{2n+1}} \left( \sum_{k=0}^{2n+1} {2n+1 \choose k} \mu^k ( x_{i+1}^{2n+1-k} - x_{i}^{2n+1-k} ) \right) \right] \right\} \\
&\times \left\{ \sum_{j=0}^N  \frac{y_j}{\sqrt{2}} \left[ \sum_{m = 0}^\infty \frac{(-1)^m}{m! (2m+1)} \frac{1}{(\sqrt{2}\sigma)^{2m+1}} \left( \sum_{l=0}^{2m+1} {2m+1 \choose l} \mu^l ( x_{j+1}^{2m+1-l} - x_{j}^{2m+1-l} ) \right) \right] \right\} ,
\end{aligned}
\end{equation}

where we have used the Maclaurin series expansion of the error function and the binomial expansion. Moving the $i, j$ summation forwards yields

@@

where $c$ is some constants that depends on the specific digitisation scheme chosen. Returning to equation \ref{eq:normcrosspower} we see
